Starcode: sequence clustering based on all-pairs search

### 2014-10-31 doc/document.tex
Documents as uploaded to Bioinformatics.
Updating the manuscript for revisions (in progress).

\title[Exact sequence clustering with starcode]{Starcode: an exact
algorithm for sequence clustering}
\author[Valera Zorita \textit{et~al}]{Eduard Valera Zorita\,$^{1,2}$, Pol
Cusc\'o\,$^{1,2}$ and Guillaume Filion\,$^{1,2}$\footnote{to whom
correspondence should be addressed}}
\address{$^{1}$Genome Architecture, Gene Regulation, Stem Cells and
Cancer Programme, Centre for Genomic Regulation (CRG), Dr. Aiguader 88,
08003 Barcelona, Spain.\\
$^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain.}

@@

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new
applications and challenges for computational biology. One such
application is the use of random barcodes to trace and quantify
transcripts or lineages in experimental setups. The high error
rate of modern sequencers calls for additional post-processing
techniques capable of detecting and reverting the misreads.
However, in the absence of a reference population, the problem
amounts to performing a pairwise comparison of all the barcodes,
which is unfeasible for excessive computationally complexity.

\section{Results:}
Here we address this problem and describe an exact algorithm to
determine which pairs of sequences lie within a given Levenshtein
distance. The matched pairs are merged into clusters represented
by a canonical sequence. The effiency of starcode is attributable
to the poucet search, a novel  implementation of the Needleman-Wunsch
algorithm performed on the nodes of a trie. On the task of clustering
random barcodes, starcode outperforms sequence clustering algorithms
in both speed and precision. We further show that starcode can also
be used to identify enriched motifs in DNA and RNA sequences.

@@

Sequence clustering is the process of grouping similar biological sequences.
It has been traditionally applied to identify related protein families and to
reduce sequence redundancy in databases. Recently, the advent of high throughput
sequencing has created additional needs for efficient clustering algorithms, in
particular because of the high error rate of such technologies. For instance,
the Illumina platform \citep{pmid16056220} shows a 1-2\% error rate consisting of
 substitutions near the 3' end of the read \citep{pmid18660515, pmid21576222}. The
PacBio platform shows a 15\% error rate that mostly corresponds to insertions
and deletions \citep{pmid19023044}. As a consequence, the same sequence is often
decoded in different ways, which artificially increases the diversity of the
output.

Sequencing errors can be discovered by mapping the reads onto a reference, if it
is available. When the sequences are random or drawn from an unknown
reference, clustering is the best option to tell real from spurious reads. One
such case is the use of random barcodes to track cells or transcripts
\citep{pmid18809713, pmid23953119}. Sequencing errors will create erroneous
barcodes that have to be reverted to the original sequence. Hands on
experience with real datasets shows that this step becomes limiting when
the number of unique sequences is high. In search for a solution to this problem, we
realized that heuristic approaches rely on assumptions that may not hold
as technologies evolve. We therefore set out to find an exact algorithm.

The first task of clustering is a matching phase where closely related barcodes are
paired, similarly to linked nodes on a graph. The second task is the clustering proper,
where communities in this graph are merged. We called our algorithm ``starcode'', in
reference to the star shape of the graph formed by the barcodes in the same
cluster. The first version of the algorithm had the same performance regardless
of the order in which the input sequences were processed, because it did not
exploit data structuring of any kind. Exploiting the prefix redundancy of
alphabetically sorted sequences allowed us to avoid unnecessary recomputations
and gain speed. This is the rationale behind the poucet search algorithm at the
heart of the matching step.

Here we describe the starcode algorithm and we benchmark it against
existing software that perform related tasks. We show that starcode is
both faster and more precise than the alternatives, achieving perfect
clustering on ideal datasets. We further show that starcode can be used
to identify enriched motifs in large datasets such as bacterial genomes
or protein-DNA/protein-RNA interaction experiments.

\begin{methods}
\section{Methods}
\subsection{Inexact string matching using tries}
The matching method of starcode is based on a variation of the Needleman-Wunsch
(NW) algorithm \citep{pmid5420325}. In the original algorithm
(Figure~\ref{fig:NW}a), the Levenshtein distance between two sequences is found
by applying a recurrence relation throughout a matrix of $mn$ terms, where $m$
and $n$ are the respective sequence lengths. The complexity of this dynamic
programming approach is $O(mn)$.

\begin{figure}[!tpb]
\centerline{\includegraphics{NW.pdf}}
\caption{Needleman-Wunsch (NW) alignment. \textbf{a}
Alignment of GTTGCA and GATCCA using the NW alignment matrix. The
margins (purple) are initialized and the cells of the matrix (yellow)
are computed from left to right and from top to bottom by the following
dynamic programming algorithm. The value in each cell is computed from
the values of the neighboring top, left and top-left cells. Denoting
$a$, $b$ and $c$ these respective terms, the value in the cell of
coordinates $(i,j)$ is computed as $\min(a+1, b+1, c)$ if the $i$-th
letter from the first sequence is identical to the $j$-th letter from
the second, or $\min(a+1, b+1, c+1)$ if the letters are different. The
Levenshtein distance between the two sequences is found in the bottom
right cell. \textbf{b} Lower complexity algorithm to determine whether
GTTGCA and GATCCA are 2-neighbors. The values in purple cells are set
during initialization. The dynamic programming algorithm proceeds as
above, with the difference that it is interrupted if the value of a
diagonal cell (bold borders) is larger than 2. The values in purple
cells are not always the same as their equivalent in \textbf{a} (purple
arrow), but the values in the yellow cells are nevertheless identical.
The values of the white cells are never computed, which contributes to
making the algorithm less complex.}\label{fig:NW}
\end{figure}

In many instances, the only information of interest is to find out
whether the sequences are $d$-neighbors (separated by a distance less
than or equal to a threshold $d$).  In that case, the complexity then
reduces to $O(d \min(m,n))$ as described below \citep{ukkonen}. Instead
of initializing the margins of the matrix and computing all the terms,
the matrix is initialized as shown on Figure~\ref{fig:NW}b and only the
terms around the diagonal are computed. If the computation of a
diagonal term yields a value greater than $d$, the distance is known to
be greater than $d$ and the process is halted. Otherwise, the
bottom-right term indicates the actual distance between the sequences,
as in the NW algorithm.

This method can be used for inexact matching of sequences against a
reference set indexed as a prefix tree, also known as a trie
\citep{ukkonen}. The terms of the matrix are updated row-wise as a
depth-first search traverses the trie from the root, as illustrated
in Figure~\ref{fig:trie}. Every time a node is visited, the row
corresponding to its depth is recomputed. If the threshold value $d$
is exceeded for a diagonal term, the Levenshtein distance for all the
downstream sequences is also necessarily greater than $d$. Therefore,
no more hits are to be discovered in this path and the depth-first
search backtracks to the parent node. When the process halts, every
tail node (corresponding to a sequence of the database) on the path
of this search is a $d$-neighbor of the query. This method is
efficient because it eliminates large areas of the search space, and
because the NW alignment of the query with each prefix of the database
is computed only once.

\begin{figure}[!tpb]
\centerline{\includegraphics{trie.pdf}}
\caption{Trie query with NW alignment. Each sequence of the index is a
path in the trie. The query GTTGCA is written at the top of a NW matrix,
which is initialized as shown on Figure~\ref{fig:NW}b. The trie is
traversed by a depth-first search (red path) from the root. At each
depth, the node added to the path is written on the left of the NW
matrix and the row is computed. Checkpoints from 1 to 4 (circled red
numbers) show the state of the NW matrix as the search proceeds. The
node labeled 3 is a leaf and thus corresponds to a 2-neighbor of the
query. The search path then backtracks to the node labeled 2
and the last rows of the NW are erased. The search path then goes to
the node labeled 4, in which case the newly computed diagonal cell
exceeds the threshold (circled in red). Even if this node has children,
they are not visited (red crosses).}
\label{fig:trie}
\end{figure}

\subsection{The poucet search algorithm}
This strategy can be
further improved. Notice that if two consecutive queries share a prefix of length
$k$, the succession of computations up to the $k$-th row of the NW matrix will be exactly
the same in both queries. To take advantage of this property, the input sequences are
sorted alphabetically in order to maximize the prefix sharing. The main idea of the
algorithm is to store the computational intermediates in the nodes of the trie and
use them later to resume the computation in the next query. This way, the search can
start at depth $k$ whenever a query shares a prefix of such length with the previous
one.

However, storing the rows of the NW matrix in the nodes, as suggested by the
approach described in the previous section is not optimal. At depth equal to the
length of the shared prefix, say $k$, the terms on the right of the diagonal were
computed using characters that belong to the previous query. Since they also depend
on the terms computed at lower depth, the search cannot restart at depth $k$. This
issue is solved by storing in each node a combination of row and column terms that
form an angle shape, looking
like a horizontally flipped L character as shown on Figure~\ref{fig:poucet}. This way, the
computation intermediates stored in a node of depth $k$ depend only on the previous
characters of the query. Using this technique, the search can effectively restart
at a depth equal to the length of the shared prefix with the previous query, thereby
averting the need to recompute the first terms of the NW alignments.

In the fairy tale ``Le Petit Poucet'', the hero seeds white pebbles for his older
brothers to find their way home, which is reminiscent of the way queries pave
 the way for the next in this algorithm. We therefore called this search
algorithm ``poucet''.

\begin{figure}[!tpb]
\centerline{\includegraphics{poucet.pdf}}
\caption{Poucet search algorithm. The algorithm proceeds with the same principles
as shown on Figure~\ref{fig:trie} with the difference that the NW matrix is not
updated row-wise, but along horizontally flipped L shapes. As the depth-first
search proceeds, these values are stored in the nodes of the trie. Only the nodes
at the top contain initialized values; for the other nodes, the values at the
border are implicitly known to be 3. Since the values in the vertical part of
the flipped L are the same for every child of the same node, they are computed
only once (purple arrow). The values in the grey cells will be computed as the
search path (red) visits the node. Storing the intermediates in the nodes allows
the next query to restart at depth $k$ if it shares a common prefix of length
$k$ with the current query.}
\label{fig:poucet}
\end{figure}

\subsection{Lookup search}
If the Levenshtein distance of two sequences of length $L$ is
less than or equal to $\tau$, there must exist a perfect match between
them of at least $k =\lfloor L / (\tau+1)\rfloor$ nucleotides. We take
advantage of this property to build a lookup table containing all such
$k$-mers of the sequences already present in the trie. When a new
sequence is queried, all its subsequences are extracted and queried
against the lookup table. If none of the subsequences are found, the
minimum distance between the query and all the sequences in the trie
is necessarily greater than $\tau$, and the trie search can therefore
be omitted. This lookup is particulary useful for long query sequences
(see \ref{performance}), where $k$ is large and the probability
of an exact match between subsequences is small unless they are
actually related, i.e. they belong to the same cluster. With this
method, the absence of match is detected at the computational cost of
several table lookups, compared to the higher cost of measuring the
Levenshtein distance between long sequences.

In general, however, $L / (\tau+1)$ is not an integer. To deal
with this complication, we divide the query into $\tau+2$ subsequences.
The first $k+1$ are called \emph{words} and have a target length
$(L-\tau) / (\tau+1)$. The last subsequence is called \emph{tail} and has
length $\tau$. The length of the words is computed by integer
division, and the remainder $R$ is distributed equally among the last
$R$ words. As a result, the words have a length that differs by at
most 1, with an aggregated length equal to $L-\tau$.

Since two exact words at different positions in the sequence shall not
represent a lookup match, we store and query the words against $k+1$
distinct lookup tables. In order to allow insertions and deletions, 
the lookup search must also query shifted versions of each word. In
the worst matching case, only one word is found and the mismatches are
evenly distributed in the query sequence. In such case, the maximum 
number of cumulative insertions/deletions at the $i$-th word is
$i-1$. Hence, the $i$-th word has to be shifted and queried against
the lookup table $2(i-1)+1$ times to cover all the possible
cases. The last word will be queried $2\tau+1$ times, including
$\tau$ shifts on the right and $\tau$ shifts on the left. For the
latter shift we make use of the $\tau$ nucleotides stored in the
tail. The total number of lookups per sequence is $(\tau+1)^2$.

\subsection{Seek and construct}
To reduce the size of the search space, we use a dynamic ``seek and construct''
approach whereby queries are processed meanwhile the trie is built. In other
words, each sequence is matched against the trie before it is inserted. To
illustrate why the trie does not need to contain all the sequences upon query,
assume that two sequences A and B are $d$-neighbors. A is processed first. Since
B is not yet inserted, A yields no hit. It is then inserted in the trie. At the
time B is processed, A is a hit for the query and the match A-B is discovered.
This approach guarantees that every hit is discovered, while maintaining the
trie as ``thin'' as possible, thereby reducing the search time.
The whole matching process is summarized in the pseudocode shown in
Algorithms~\ref{alg:starcode} and \ref{alg:poucet}.

@@

\subsection{Parallelization}
To parallelize the search, queries are separated into contiguous blocks after
sorting. The matching algorithm proceeds in two phases. During the
first phase, an independent trie is created and filled with the
sequences of its associated block using the seek and construct process
described above. In the second phase, all the other blocks of
sequences are queried against but not inserted into each trie built in
the first phase. If the queries are segregated into $N$ blocks, the
first phase consists of $N$ seek and construct jobs, whereas the second
consists of $N(N-1)/2$ query jobs. Since the jobs show little
dependence on each other, the matching algorithm can be efficiently
parallelized provided $N$ is larger than the number of independent
threads.
\subsection{Clustering}
Starcode implements a multi-purpose clustering algorithm called ``sphere
clustering'' \citep{pmid23953119}, and a message passing algorithm
\citep{mackay} tailored for the task of clustering random barcodes. In
sphere clustering, barcodes are sorted by frequency of occurrence and
each barcode, starting from the most frequent, can claim its
$d$-neighbors that were not already claimed.

In message passing clustering, read counts are distributed equally
among the closest neighbors of each barcode only if they are at least
5 times more frequent. The remaining sequences at the end of the
process are considered canonical, and their associated count is the
estimated cluster size. A barcode is assigned to a cluster if all its
read counts are eventually given to the corresponding canonical
barcode. The barcodes for which the read counts are split between
different canonical barcodes are not assigned to any cluster. The
reason for imposing a factor 5 or larger in order to transfer the read
counts is that barcodes with similar frequencies are not likely
derived from each other through sequencing errors. More likely they
are either unrelated, or they both derive from another more abundant
barcode.

@@

\subsection{Performance}
\label{performance}
We measured basic performance and scalability metrics of starcode on a
dataset of pseudo random sequences (Figure~\ref{fig:perf}). The standard
configuration consists of a set of 1,000,000 sequences of length 40 running
on 1 thread and with a maximum Levenshtein distance of 3. To test the
scalability as a function of a single parameter, only the parameter under
study was modified whereas the others were kept constant.

Figure~\ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic scale
the trend is a straight line with slope 1.7, which suggests that the
running time complexity of starcode is approximately $O(n^{1.7})$.
Figure~\ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
This is a common feature of exact algorithms based on trie search. As
this parameter increases, the poucet search bails out at a deeper
average depth in a trie that fans out exponentially.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure~\ref{fig:perf}c).
Beyond a certain length, the lookup search algorithm starts to be
efficient, and most of the queries are resolved without the need for
trie-search.
Finally, we show the scalability of starcode with increasing
number of threads in Figure~\ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 8 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is that there is insufficient memory bandwidth to satisfy
the increased demand of memory accesses due to hardware limitations.

\begin{figure}[!tpb]%figurex
\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
\caption{Scalability. a. Logarithm of the running time versus the
logarithm of the number of sequences to be clustered. b. Running time as
a function of the clustering distance. c. Running time versus length of
the input sequences. d. Relative performance increase for different number
of parallel threads. 
}\label{fig:perf}
\end{figure}

\subsection{Benchmark against sequence clustering algorithms}
Sequence clustering is routinely used to curate databases of non
redundant nucleotide or protein sequences. We benchmarked starcode
against the two popular sequence clustering algorithms CD-HIT
\citep{pmid23060610} and USEARCH \citep{pmid20709691}.

We generated 1 million random 40-mers duplicated 47 times. For each
40-mer, we also inserted 3 additional mutant sequences to the pool
by sampling again 3 nucleotides at random positions. The probability
that such a mutant sequence is within a distance 3 of any other
40-mer is of the order of $10^{-12}$ and can be neglected. Each
cluster thus consists of 50 sequences, of which 3 have a Levenshtein
distance 3 from the canonical representant (referred to as the centroid
in clustering terms).

All software were set to run on a 6-core single-processor Intel Xeon
E5-2620 system with 48~GB of DDR3-RAM at 1333~Mhz.
Using 12 threads, starcode clustered the shuffled 50 million sequences
in 5 minutes and 3 seconds without a single error (the output
consisted of 1 million clusters of size 50). There is no obvious
way to compare exact algorithms with heuristic algorithms such as
CD-HIT and USEARCH. We decided to allocate the heuristics 6 minutes
on the same dataset (20\% additional time) and use the
parameters giving the best accurary under this constraint.
However, we could not find a combination of
parameters such that either of them could terminate in less than
6 minutes. To simplify the task, we extracted the unique sequences
from the input file, bringing the number down to 4,000,000 and followed
the same rationale. We could not find any set of parameters such
that USEARCH could process this input in less than 6 minutes
(because of poor multi-threading performance), but CD-HIT could
perform the task when run as shown in Table~\ref{tab:params1}.
In such conditions, it identified 2,798,656 clusters, indicating that
the clustering was not nearly complete.

>>>>

\title[Sequence clustering by all pairs-search]{Starcode: sequence
clustering based on all-pairs search }
\author[Valera Zorita \textit{et~al}.]{Eduard Valera Zorita\,$^{1,2}$, Pol
Cusc\'o\,$^{1,2}$ and Guillaume Filion\,$^{1,2}$\footnote{to whom
correspondence should be addressed}}
\address{$^{1}$Genome Architecture, Gene Regulation, Stem Cells and
Cancer Programme, Centre for Genomic Regulation (CRG), Dr. Aiguader 88,
08003 Barcelona, Spain.\\
$^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain.}

@@

\begin{abstract}
\section{Motivation:}
The increasing throughput of sequencing technologies offers new
applications and challenges for computational biology. In many
of those applications, sequencing errors need to be corrected.
This is particularly important when using random DNA barcodes to
trace and quantify transcripts or lineages, since spurious barcodes
introduce inexistent entities. The reference population of random
barcodes is unknown and in this case, error correction amounts
to performing a pairwise comparison of all the barcodes,
which is unfeasible for excessive computationally complexity.

\section{Results:}
Here we address this problem and describe an exact algorithm to
determine which pairs of sequences lie within a given Levenshtein
distance. For error correction or redundancy reduction purposes,
matched pairs are then merged into clusters represented by a
canonical sequence. The effiency of starcode is attributable to
the poucet search, a novel implementation of the Needleman-Wunsch
algorithm performed on the nodes of a trie. On the task of matching
random barcodes, starcode outperforms sequence clustering algorithms
in both speed and precision.

@@

Sequencing technologies never achieve perfect precision. For instance,
the Illumina platform \citep{pmid16056220} has a 1-2\% error rate
consisting of substitutions \citep{pmid18660515, pmid21576222}
and the PacBio platform has a 15\% error rate consisting of insertions
and deletions \citep{pmid19023044}. The throughput of such technologies
has recently created additional needs to develop efficient error
correction algorithms.

Sequencing errors can be discovered by comparing the reads to a
reference genome. However, such a reference is not always available.
When the sequences are random or taken from an unknown source,
clustering is the main strategy to correct the errors.
For instance, this situation arises when using random barcodes to
track cells or transcripts \citep{pmid18809713, pmid23953119}.
Sequencing errors will create erroneous (nonexistent) barcodes
that have to be removed.

Sequence clustering can be viewed as a community detection problem
on graphs, where nodes represent sequences and edges represent matches
between related sequences. The process consists of a matching phase
(the most computationally intensive), where the graph is constructed,
and a clustering phase where communities are identified.

Here we describe a sequence clustering algorithm focused on error
correction. The algorithm is called ``starcode'' in reference to 
clusters of random barcodes, which typically have a star shape.
Starcode is based on all-pairs search, \textit{i.e.} all the pairs
of sequences below a given Levenshtein distance are identified during
the graph construction phase. Matching is carried out by lossless
filtration, followed by a trie search.  The novelty of the algorithm
is the poucet strategy, which uses the redundancy of alphabetically
sorted sequences to avoid unnecessary recomputations and gain speed.

In this article we present and benchmark starcode. We show that on
real biological datasets, starcode is faster than existing sequence
clustering software. Even though starcode was designed for error
correction, we show that it can be used for other problems. As an
illustration, we use it to identify enriched motifs in a bacterial
genome and in protein-RNA interaction experiments.

\begin{methods}
\section{Methods}
\subsection{Inexact string matching using tries}
The matching method of starcode is based on a variation of the
Needleman-Wunsch (NW) algorithm \citep{pmid5420325}. In the
original algorithm (Figure~\ref{fig:NW}a), the Levenshtein
distance between two sequences is found by applying a recurrence
relation throughout a matrix of $mn$ terms (the edit matrix),
where $m$ and $n$ are the respective sequence lengths. The complexity
of this dynamic programming approach is $O(mn)$.

\begin{figure}[!tpb]
\centerline{\includegraphics{NW.pdf}}
\caption{Needleman-Wunsch (NW) sequence comparison. \textbf{a}
Comparison of GTTGCA and GATCCA. The margins of the edit matrix
(purple) are initialized and the cells (yellow)
are computed from left to right and from top to bottom by the NW
dynamic programming algorithm. $E[i,j]$, the term of
coordinates $(i,j)$ is computed as
$\min(E[i-1,j]+1, E[i,j-1]+1, E[i-1,j-1]+\Delta(i,j))$, where
$\Delta(i,j) = 0$ if the $i$-th symbol from the first sequence is the
same as the $j$-th symbol from the second, and $\Delta(i,j) = 1$
otherwise. The Levenshtein distance between the two sequences is the
value of the bottom right cell. \textbf{b} Lower complexity algorithm
to determine whether GTTGCA and GATCCA are 2-matches.
The values in the purple cells are set
during initialization. The dynamic programming algorithm proceeds as
above, with the difference that it is aborted if the value of a
diagonal cell (bold borders) is larger than 2. The values in the
purple cells may differ from the original NW scheme (purple arrow),
but the values in the yellow cells are nevertheless identical.
The values of the white cells are never computed, which contributes to
reducing the complexity.}\label{fig:NW}
\end{figure}

In many instances, the information of interest is to find out
whether the sequences are $\tau$-matches (\textit{i.e.} their distance
is less then or equal to a fixed threshold $\tau$).
In that case, the complexity can be reduced to $O(\tau \min(m,n))$.
Instead of computing all the terms of the edit matrix, it is initialized
as shown on Figure~\ref{fig:NW}b and only the terms around the diagonal
are computed. If a diagonal term has a value greater than $\tau$,
the process is halted because the sequences are not $\tau$-matches.

This method can be used to match sequences against a prefix tree,
also known as a trie \citep{ukkonen}. The terms of the edit matrix
are updated row-wise while a depth-first search traverses the trie
(Figure~\ref{fig:trie}). Every time a node is visited,
a row is computed, and every time the search backtracks, a row is
erased. If the threshold value $\tau$
is exceeded for a diagonal term, the Levenshtein distance for all the
downstream sequences is also necessarily greater than $\tau$. Therefore,
no more hits are to be discovered in this path and the depth-first
search backtracks to the parent node. When the process halts, every
tail node (corresponding to a sequence of the database) on the path
of this search is a $\tau$-match of the query. This method is
efficient because it eliminates large areas of the search space, and
because the NW comparison of the query with each prefix of the database
is computed only once.

\begin{figure}[!tpb]
\centerline{\includegraphics{trie.pdf}}
\caption{NW algorithm on tries. Each sequence of the index is a
path in the trie. The query GTTGCA is written at the top of the matrix,
which is initialized as shown on Figure~\ref{fig:NW}b. The trie is
traversed by a depth-first search (red path). At each
depth, the node added to the path is written on the left of the edit
matrix and the row is computed. Checkpoints from 1 to 4 (circled red
numbers) show the state of the edit matrix as the search proceeds. The
node labeled 3 is a leaf and thus corresponds to a 2-match of the
query. After discovering the hit, the search path backtracks to the
node labeled 2 and the last rows of the edit matrix are erased.
The search path then goes to the node labeled 4, in which case the
newly computed diagonal cell exceeds the threshold (circled in red).
Even if this node has children, they are not visited (red crosses)
because there is no 2-match to discover.}
\label{fig:trie}
\end{figure}

\subsection{The poucet search algorithm}
The search strategy can be further improved. If two consecutive
queries share a prefix of length $k$, the succession of computations
up to the $k$-th row of the edit matrix will be exactly the same
for both queries. Therefore, computation intermediates can be
stored in the nodes of the trie, so that the next trie search can start
at depth $k$. However, storing the rows of the edit matrix in the nodes
meets some difficulty. Indeed, on the $k$-th row, the terms on the
right side of the diagonal depend on characters that are not shared
between the two queries. This issue is solved by storing in each
node a combination of row and column terms that form an angle shape,
looking like a horizontally flipped L (Figure~\ref{fig:poucet}).
Using this structure, the computation intermediates stored in a
node at depth $k$ depend only on the first $k$ characters of the
query.

To take full advantage of this property, the input sequences are
sorted alphabetically, which maximize prefix sharing between
consecutive queries. In the fairy tale ``Le Petit Poucet'', the
hero seeds white pebbles for his older brothers to find their way
home, which is reminiscent of the way a smaller query (in alphabetical
order) paves the way for the next. We therefore called this search
algorithm ``poucet''.

\begin{figure}[!tpb]
\centerline{\includegraphics{poucet.pdf}}
\caption{Poucet search algorithm. The algorithm proceeds with the
same principles as shown on Figure~\ref{fig:trie} with the difference
that the edit matrix is not updated row-wise, but along a horizontally
flipped L. As the depth-first search proceeds, these values
are stored in the nodes of the trie. Since the values in the vertical
part of the flipped L are the same for every child of a node,
they are computed only once (purple arrow). The values in the grey
cells will be computed as the search path (red) visits the node.
Storing the intermediates in the nodes allows the next query to
restart at depth $k$ if it shares a common prefix of length
$k$ with the current query.}
\label{fig:poucet}
\end{figure}

\subsection{Lossless filtration}
When a query has no match, it is advantageous to omit the trie
search. To this end, starcode uses a partition
approach similar to that described by \cite{WuManber1992}. If the
query is partitioned in $\tau+1$ segments, every $\tau$-match
contains at least one of those (assuming that all the segments
have length at least $\tau$). Indeed, there are at most $\tau$
differences between the query and the match, so at least one segment
is unmodified. In the $\tau$-match, the position of the shared
segment is at most $\tau$ nucleotides on the left or on the right
of its position in the query.

These observations are the basis of a filtration method that proceeds
as follows. Every time a sequence is added to the trie, its segments
are added to $\tau+1$ different indexes. More specifically, the
first $\tau$ nucleotides of the sequence are removed, and the rest
of the sequence is partitioned in $\tau+1$ segments of sizes
differing by at most 1 (the longer segments always in 3' for
consistency). The first fragments are added to the first index,
the second fragments to the second index \textit{etc}.

Before the search, segments of the query are looked up in
the indexes, and in case no match is found, the trie search is
omitted because this query has no $\tau$-match. If at least one
segment is found, the trie search is performed. As mentioned above,
segments shared between the query and a $\tau$-match can be shifted
up to $\tau$ nucleotides. For this reason, shifted segments are
looked up in the indexes according to the following scheme ensuring
that no match can be missed.
(Figure~\ref{fig:filtration}). The rightmost segment is looked up
in the $\tau$+1-th index, the second rightmost segment and the
contiguous segments shifted by 1 nucleotides are are looked up in the
$\tau$-th index and so on until the first segment and its
contiguous segments shifted by up to $\tau$ nucleotides are
looked up in the first index.

\begin{figure}[!tpb]
\centerline{\includegraphics{filtration.pdf}}
\caption{Lossless filtration illustrated by an example sequence
of length 20 with $\tau=3$. The last $\tau$ nucleotides of the query
are removed, and the rest is divided into 4 series of contiguous
segments.  Each series is queried against a different index numbered
I to IV.  For instance, the only segment queried against index I
is GTTG, while those queried against index II are GCAA, CAAT and AATA.
If any of the segments is found in the appropriate index, the trie
search is performed, otherwise it is omited as there can be no
$\tau$-match. Regardless of the result, segments labelled I to IV
are then added to the corresponding respective index (\textit{i.e.}
only one segment is added to each index).}
\label{fig:filtration}
\end{figure}

\subsection{Seek and construct}
To reduce the size of the search space, starcode uses a dynamic
``seek and construct'' approach whereby queries are processed
meanwhile the trie is built. In other words, each sequence is
matched against the trie before it is inserted. If A and B are
mutual $\tau$-matches, either A will be queried when B is in
in the trie, or the converse. Either way, the match A-B is
discovered. This guarantees that every $\tau$-match is discovered,
while maintaining the trie as ``thin'' as possible, thereby reducing
the search time.  The whole matching process is summarized in
the pseudocode shown in Algorithms~\ref{alg:starcode} and
\ref{alg:poucet}.

@@

\subsection{Parallelization}
Queries are sorted and partitioned in contiguous blocks.
The matching step then proceeds in two phases. In the first, a
distinct trie is built from the sequences of each block according
to the algorithm described above. In the second, all the sequence
blocks are queried against the tries. If the queries are partitioned
in $N$ blocks, the first phase consists of $N$ seek and construct
jobs, wile the second consists of $N(N-1)/2$ query jobs. In each
phase, the jobs show little dependency on each other, so the
matching algorithm can be efficiently parallelized provided $N$
is larger than the number of independent threads.

\subsection{Clustering}
The default clustering algorithm of starcode is designed to correct
sequencing error. This method uses message passing \citep{mackay}
to identfy and count ``canonical'' sequences (also referred to as
centroids in the clustering terminology). Each sequence transfers its
read count to its closest $\tau$-match provided the latter has at least 5
times more counts. If the condition is not met, the transfer does
not take place. If the sequence has several equally close
$\tau$-matches, the counts are split equally among them. The
process is repeated recursively, starting from sequences with
lowest read count. The sequences with a positive read count at
the end of the process are considered canonical. Clusters consist
of all the sequences transferring their read counts to the same
canonical sequence (sequence transferring their read counts to
different canonicals are discarded).

Since no sequencing technology has an error rate higher than 20\%,
it is expected that sequences appearing from sequencing errors will
always have 5 times or lower read count than the canonical sequence.
Otherwise, sequences are more likely unrelated, or both derived from
the same canonical sequence.

For other sequence clustering problems, starcode implements a
multi-purpose algorithm called ``sphere clustering''
\citep{pmid23953119}. In sphere clustering, sequences are sorted by
frequency of occurrence. Starting from the most
frequent, each sequence becomes canonical and claims all its
$\tau$-matches, which forms a cluster of radius $\tau$ (hence the
name). Claimed sequences are immediately removed, so that they can
belong to only one cluster.

\subsection{Benchmark conditions}
All the tests were performed on a 16-core dual-processor
Intel Xeon E5-2687W v2 system with 256~GB of DDR3-RAM at 1866~Mhz

@@

\subsection{Basic performance}
\label{performance}
We measured basic performance and scalability metrics of starcode on a
dataset of pseudo random sequences (Figure~\ref{fig:perf}). The standard
configuration consists of a set of 1,000,000 sequences of length 40 running
on 1 thread and with a maximum Levenshtein distance of 3. To test the
scalability as a function of a single parameter, only the parameter under
study was modified whereas the others were kept constant.

Figure~\ref{fig:perf}a shows the running time of starcode as a
function of the number of input sequences $n$. In double logarithmic scale
the trend is a straight line with slope 1.5, suggesting that the
running time complexity of starcode is lower than quadratic (unlike
the naive implementation of all-pairs search).
Figure~\ref{fig:perf}b shows that the running time grows exponentially
as a function of the maximum Levenshtein distance used for clustering.
The reason is that the trie fans out exponentially and the search bails
out at a greater depth as the maximum distance increases.
As a function of the sequence length, the running time first increases 
but then plummets and stays low (Figure~\ref{fig:perf}c). Beyond a
threshold length, the filtering algorithm starts to be efficient, and
most of the queries are resolved without searching the trie.
Finally, we show the scalability of starcode with increasing
number of threads in Figure~\ref{fig:perf}d. The search algorithm is fully
parallel and the relative performance increases linearly up to 12 threads.
The bending observed thereafter has two sources. The first is 
that the input reading and clustering steps are brief but not parallel,
the second is that there is insufficient memory bandwidth to satisfy
the increased demand of memory accesses due to hardware limitations.

\begin{figure}[!tpb]
% Basic performance figure.
\centerline{\includegraphics[scale=0.47]{scalability.pdf}}
\caption{Scalability. \textbf{a} Logarithm of the running time versus the
logarithm of the number of sequences to be clustered. \textbf{b}
Running time as a function of the clustering distance. \textbf{c}
Running time versus length of the input sequences. \textbf{d} Relative
performance increase for different number of parallel threads. 
}\label{fig:perf}
\end{figure}

\subsection{Benchmark}
We benchmarked starcode against the sequence clustering algorithms
slidesort \citep{pmid21148542}, seed \citep{pmid21810899},
rainbow \citep{pmid22942077} and cd-hit \citep{pmid23060610}. Even though
slidesort is an all-pairs search algorithm, it was included in the
benchmark because sequence comparison is the most computationally
intensive step of the sequence clustering problem. Rainbow runs
exclusively on paired-end reads, while the other tools run on single
reads.

%% Synthetic data %%
The performance of sequence clustering algorithms can be sensitive
to the size of the clusters in the dataset, which in many applications
is not known \textit{a priori}. We therefore set up a benchmark on
artificial datasets to test the accuracy and the scaling of the
tools on a known cluster structure. We generated 4 datasets of 1 million
50-mers arranged in 1 to 1,000 clusters. Each cluster consisted
of 100 repeats of the same centroid sequence, plus satellites
derived from the centroid by incorporating 3 errors including at
most 1 indel. The number of satellites per cluster ranged from
999900 to 900. Evaluating the exactness of slidesort on these datasets
was problematic because the number of 3-matches in each dataset is
not known (pairs of statellites in the same cluster may be 3-matches
or not). For this reason we only compared the number of pairs found
by starcode with the number of pairs found by slidesort. The 
outcome of the test is summarized in
Figure~\ref{fig:benchmark}.

\begin{figure}[!tpb]
  \centerline{\includegraphics[scale=0.47]{benchmark.pdf}}
\caption{Accuracy of the sequence clustering and all-pairs search
algorithms. Starcode, seed and cd-hit were bencmarked on 4 datasets
with the same number of sequences, but different number of clusters
(see main text). \textbf{a} Starcode identifies the correct number
of clusters, while seed and cd-hit identify about 40 false positives
per true positive. \textbf{b} Slidesort identifies 5-10\% less
pairs than starcode. In both panels, the red line is first
bisector.}
\label{fig:benchmark}
\end{figure}

While starcode achieves perfect clustering on all 4 datasets, the
clustering achieved by seed and cd-hit is incomplete. Both tools
identify approximately 40 false clusters per true cluster on all
the datasets. We also observed that slidesort found 5-10\% less
3-matches than starcode on all the datasets. We were surprised by
this result because slidesort is claimed to be an exact algorithm.
However, this was clearly not the case when we ran additional tests
on smaller datasets where naive pairwise comparisons is feasible
(the starcode repository contains such a dataset along with the
instructions to reproduce our observations, see
http://github.com/gui11aume/starcode/misc). In conclusion, starcode
was the only tool to achieve perfect precision.
